---
title: "Economic Policy Attitudes 2PL IRT Response Simulation"
---

# Economic Policy Attitudes — 2PL IRT Simulation & Parameter Recovery

Load libraries. 

```{r}
library(mirt)
library(ggplot2)
library(dplyr)
library(tidyr)

set.seed(2024)  # For reproducibility
```

This script:
1. Assigns provisional a (discrimination) and b (difficulty/location) parameters to all 30 items based on content review
2. Simulates 500 respondents using the 2PL model
3. Exports the response matrix as a CSV
4. Fits a 2PL model to the simulated data using mirt
5. Compares recovered parameters to generating ("true") parameters
6. Produces ICC plots and a parameter recovery summary

SCORING CONVENTION:
0 = Progressive response (lower theta)
1 = Conservative response (higher theta)

THETA SCALE: Standard normal N(0,1), range approx -3 to +3
```{r}

# =============================================================================
# SECTION 1: Provisional Item Parameters
# =============================================================================
# a = discrimination  (how steeply the item separates respondents)
#     High disc items:     a ~ 1.5-2.5
#     Moderate disc items: a ~ 0.8-1.4
#     Low disc items:      a ~ 0.4-0.7
#
# b = item location on the theta continuum (difficulty)
#     Negative b: item "tips" conservative at lower/more progressive theta
#     Positive b: item "tips" conservative only at higher/more conservative theta
#     b ~ 0: item is roughly 50/50 near the center of the distribution
#
# These are informed provisional values based on item content review.
# The simulation will let you verify that mirt recovers them well.
# =============================================================================

item_params <- data.frame(
  item_id   = 1:30,
  item_name = c(
    "Tax bracket >$1M",          # 1
    "Road repair funding",        # 2
    "Estate/inheritance tax",     # 3
    "Income vs. sales tax",       # 4
    "Wealth tax",                 # 5
    "Deficit spending",           # 6
    "Recession stimulus",         # 7
    "Public transit funding",     # 8
    "Defense vs. social cuts",    # 9
    "Housing subsidies",          # 10
    "Minimum wage increase",      # 11
    "Union organizing rights",    # 12
    "Paid family leave mandate",  # 13
    "Right-to-work laws",         # 14
    "Executive pay ratios",       # 15
    "Single-payer healthcare",    # 16
    "Drug price negotiation",     # 17
    "Free public college",        # 18
    "Rural broadband access",     # 19
    "Public library expansion",   # 20
    "Big tech regulation",        # 21
    "Worker board representation",# 22
    "Environmental regulations",  # 23
    "Corporate disclosure",       # 24
    "Financial institution rules",# 25
    "Tariffs on imports",         # 26
    "Trade agreement conditions", # 27
    "Foreign acquisition review", # 28
    "Buy American requirement",   # 29
    "Trade-displaced worker aid"  # 30
  ),
  category = c(
    rep("Taxation & Wealth Inequality", 5),
    rep("Government Spending & Debt", 5),
    rep("Labor Markets & Worker Rights", 5),
    rep("Healthcare & Public Services", 5),
    rep("Corporate Regulation", 5),
    rep("Trade & Economic Nationalism", 5)
  ),
  disc_label = c(
    "High","Moderate","High","Low","Moderate",        # Items 1-5
    "High","Low","Moderate","Moderate","Low",          # Items 6-10
    "High","Moderate","Low","Moderate","High",         # Items 11-15
    "High","Moderate","Moderate","Low","Low",          # Items 16-20
    "High","Moderate","Moderate","Low","High",         # Items 21-25
    "High","Moderate","Moderate","Low","Low"           # Items 26-30
  ),
  # True generating discrimination parameters
  a_true = c(
    2.0,  1.2,  2.2,  0.6,  1.1,   # Taxation
    2.1,  0.5,  1.0,  1.0,  0.6,   # Gov spending
    2.3,  1.3,  0.5,  1.1,  1.8,   # Labor
    2.4,  1.2,  1.1,  0.5,  0.4,   # Healthcare
    1.9,  1.1,  1.0,  0.6,  2.0,   # Corp regulation
    1.8,  1.2,  1.0,  0.6,  0.5    # Trade
  ),
  # True generating location/difficulty parameters
  # Negative = item location on progressive side of continuum
  # Positive = item location on conservative side of continuum
  b_true = c(
    -1.2,  0.1, -1.5,  0.2, -0.8,  # Taxation: most items lean progressive
     0.3,  0.0,  0.2,  0.1,  0.3,  # Gov spending: near center
    -1.0, -0.5,  0.3, -0.2, -0.7,  # Labor: lean slightly progressive
    -1.4, -0.4, -0.3,  0.2,  0.4,  # Healthcare: lean progressive
    -0.6,  0.0,  0.3,  0.5, -0.5,  # Corp regulation: mixed
     0.5,  0.3,  0.4,  0.6,  0.2   # Trade: lean slightly conservative
  ),
  stringsAsFactors = FALSE
)

cat("=== Provisional Item Parameters ===\n")
print(item_params[, c("item_id","item_name","disc_label","a_true","b_true")])

```

## Simulate 500 Respondents
```{r}
# =============================================================================
# SECTION 2: Simulate 500 Respondents
# =============================================================================
# Theta drawn from N(0,1) — standard IRT scale.
# Using a slightly heavy-tailed distribution (t with df=10) to ensure
# adequate coverage of extreme items near |b| > 1.5.

n_persons <- 500
n_items   <- 30

# Simulate theta: mix of N(0,1) and slight heavy tails
theta_true <- rt(n_persons, df = 10)
theta_true <- theta_true / sd(theta_true)  # Rescale to unit variance

cat("\n=== Theta Distribution Summary ===\n")
cat(sprintf("Mean: %.3f | SD: %.3f | Min: %.3f | Max: %.3f\n",
            mean(theta_true), sd(theta_true),
            min(theta_true), max(theta_true)))

# 2PL probability function
p_2pl <- function(theta, a, b) {
  1 / (1 + exp(-a * (theta - b)))
}

# Generate response matrix
# Rows = persons, Columns = items
response_matrix <- matrix(NA, nrow = n_persons, ncol = n_items)

for (j in 1:n_items) {
  probs <- p_2pl(theta_true, item_params$a_true[j], item_params$b_true[j])
  response_matrix[, j] <- rbinom(n_persons, size = 1, prob = probs)
}

colnames(response_matrix) <- paste0("item_", sprintf("%02d", 1:n_items))
rownames(response_matrix) <- paste0("person_", 1:n_persons)

cat("\n=== Response Matrix Dimensions ===\n")
cat(sprintf("%d persons x %d items\n", nrow(response_matrix), ncol(response_matrix)))
cat(sprintf("Overall proportion conservative (scored 1): %.3f\n",
            mean(response_matrix)))

# Item-level endorsement rates
endorsement <- colMeans(response_matrix)
cat("\n=== Item Endorsement Rates (proportion conservative) ===\n")
print(round(endorsement, 3))

```

```{r}
# =============================================================================
# SECTION 3: Export Response Matrix
# =============================================================================

response_df <- as.data.frame(response_matrix)
response_df$theta_true <- round(theta_true, 4)

write.csv(response_df,
          "outputs/simulated_responses_500.csv",
          row.names = TRUE)

cat("\n=== Exported: simulated_responses_500.csv ===\n")
cat("Columns: item_01 through item_30, plus theta_true\n")
cat("Rows: 500 simulated respondents\n")


```

```{r}
# =============================================================================
# SECTION 4: Fit 2PL Model with mirt
# =============================================================================

cat("\n=== Fitting 2PL Model ===\n")

fit_2pl <- mirt(
  data      = as.data.frame(response_matrix),
  model     = 1,              # Unidimensional
  itemtype  = "2PL",
  SE        = TRUE,           # Compute standard errors
  verbose   = FALSE
)

cat("\n=== Model Fit Summary ===\n")
print(summary(fit_2pl))

# Global fit statistics
cat("\n=== Global Fit Statistics ===\n")
tryCatch({
  fit_stats <- M2(fit_2pl)
  print(fit_stats)
  cat("\nInterpretation:\n")
  cat("  RMSEA < 0.05 = good fit\n")
  cat("  SRMSR < 0.08 = acceptable fit\n")
}, error = function(e) {
  cat("M2 statistic not available:", e$message, "\n")
  # Fallback: -2LL and AIC/BIC
  cat("AIC:", AIC(fit_2pl), "\n")
  cat("BIC:", BIC(fit_2pl), "\n")
})
```

```{r}
# =============================================================================
# SECTION 5: Parameter Recovery Comparison
# =============================================================================

# Extract estimated parameters
# mirt parameterizes 2PL as: a1 (slope) and d (intercept, where d = -a*b)
# We convert back to a, b for comparison
params_estimated <- coef(fit_2pl, IRTpars = TRUE, simplify = TRUE)$items

# Build recovery comparison table
recovery <- data.frame(
  item_id    = item_params$item_id,
  item_name  = item_params$item_name,
  category   = item_params$category,
  disc_label = item_params$disc_label,

  # True generating parameters
  a_true = item_params$a_true,
  b_true = item_params$b_true,

  # Recovered parameters
  a_est  = round(params_estimated[, "a"], 3),
  b_est  = round(params_estimated[, "b"], 3),

  stringsAsFactors = FALSE
)

# Compute recovery statistics
recovery$a_bias  <- round(recovery$a_est - recovery$a_true, 3)
recovery$b_bias  <- round(recovery$b_est - recovery$b_true, 3)
recovery$a_abias <- abs(recovery$a_bias)
recovery$b_abias <- abs(recovery$b_bias)

cat("\n=== Parameter Recovery Table ===\n")
print(recovery[, c("item_id","item_name","a_true","a_est","a_bias",
                   "b_true","b_est","b_bias")])

# Summary statistics
cat("\n=== Recovery Summary Statistics ===\n")
cat(sprintf("Discrimination (a):\n"))
cat(sprintf("  Mean bias:  %+.4f\n", mean(recovery$a_bias)))
cat(sprintf("  Mean |bias|: %.4f\n", mean(recovery$a_abias)))
cat(sprintf("  RMSE:        %.4f\n", sqrt(mean(recovery$a_bias^2))))
cat(sprintf("  Correlation (true vs. est): %.4f\n",
            cor(recovery$a_true, recovery$a_est)))

cat(sprintf("\nLocation (b):\n"))
cat(sprintf("  Mean bias:  %+.4f\n", mean(recovery$b_bias)))
cat(sprintf("  Mean |bias|: %.4f\n", mean(recovery$b_abias)))
cat(sprintf("  RMSE:        %.4f\n", sqrt(mean(recovery$b_bias^2))))
cat(sprintf("  Correlation (true vs. est): %.4f\n",
            cor(recovery$b_true, recovery$b_est)))

# Export recovery table
write.csv(recovery, "outputs/parameter_recovery.csv", row.names = FALSE)
cat("\nExported: parameter_recovery.csv\n")
```


```{r}
# =============================================================================
# SECTION 6: Item-Level Fit Statistics
# =============================================================================

cat("\n=== Item-Level Fit (S-X2 statistic) ===\n")
tryCatch({
  item_fit <- itemfit(fit_2pl, fit_stats = "S_X2")
  print(item_fit)
  cat("\nNote: p < .05 flags items that may not fit the 2PL model well.\n")
}, error = function(e) {
  cat("Item fit statistics:", e$message, "\n")
  item_fit_alt <- itemfit(fit_2pl)
  print(item_fit_alt)
})

# =============================================================================
# SECTION 7: Theta Estimation & Recovery
# =============================================================================

cat("\n=== Theta Recovery ===\n")

# Estimate theta for all respondents using EAP (Expected A Posteriori)
theta_eap <- fscores(fit_2pl, method = "EAP", full.scores = TRUE)
theta_est <- as.vector(theta_eap[, 1])

theta_corr <- cor(theta_true, theta_est)
theta_rmse <- sqrt(mean((theta_true - theta_est)^2))

cat(sprintf("EAP theta recovery:\n"))
cat(sprintf("  Correlation (true vs. EAP): %.4f\n", theta_corr))
cat(sprintf("  RMSE:                       %.4f\n", theta_rmse))
cat(sprintf("  (r > 0.90 indicates adequate recovery with 30 items)\n"))

# =============================================================================
# SECTION 8: Plots (ggplot2)
# =============================================================================

theta_seq <- seq(-4, 4, length.out = 300)

cat_colors <- c(
  "Taxation & Wealth Inequality"     = "#185FA5",
  "Government Spending & Debt"       = "#0F6E56",
  "Labor Markets & Worker Rights"    = "#854F0B",
  "Healthcare & Public Services"     = "#993556",
  "Corporate Regulation"             = "#533AB7",
  "Trade & Economic Nationalism"     = "#A32D2D"
)

# --- Plot 1: Parameter Recovery Scatter Plots ---

a_label <- sprintf("r = %.3f  |  RMSE = %.3f",
                   cor(recovery$a_true, recovery$a_est),
                   sqrt(mean(recovery$a_bias^2)))
b_label <- sprintf("r = %.3f  |  RMSE = %.3f",
                   cor(recovery$b_true, recovery$b_est),
                   sqrt(mean(recovery$b_bias^2)))
t_label <- sprintf("r = %.3f  |  RMSE = %.3f", theta_corr, theta_rmse)

p_a <- ggplot(recovery, aes(x = a_true, y = a_est, color = category)) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "gray60") +
  geom_smooth(aes(group = 1), method = "lm", se = FALSE,
              color = "#A32D2D", linewidth = 0.8) +
  geom_point(size = 3, alpha = 0.85) +
  scale_color_manual(values = cat_colors, name = NULL) +
  annotate("text", x = -Inf, y = Inf, hjust = -0.1, vjust = 1.5,
           label = a_label, size = 3.5, color = "gray30") +
  labs(title = "Discrimination Parameter Recovery (a)",
       x = "True a", y = "Estimated a") +
  theme_minimal(base_size = 12) +
  theme(legend.position = "bottom",
        legend.text = element_text(size = 8),
        plot.title = element_text(face = "bold", size = 13))

p_b <- ggplot(recovery, aes(x = b_true, y = b_est, color = category)) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "gray60") +
  geom_smooth(aes(group = 1), method = "lm", se = FALSE,
              color = "#A32D2D", linewidth = 0.8) +
  geom_point(size = 3, alpha = 0.85) +
  scale_color_manual(values = cat_colors, name = NULL) +
  annotate("text", x = -Inf, y = Inf, hjust = -0.1, vjust = 1.5,
           label = b_label, size = 3.5, color = "gray30") +
  labs(title = "Item Location Parameter Recovery (b)",
       x = "True b", y = "Estimated b") +
  theme_minimal(base_size = 12) +
  theme(legend.position = "bottom",
        legend.text = element_text(size = 8),
        plot.title = element_text(face = "bold", size = 13))

theta_df <- data.frame(theta_true = theta_true, theta_est = theta_est)

p_theta <- ggplot(theta_df, aes(x = theta_true, y = theta_est)) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "gray60") +
  geom_point(alpha = 0.25, color = "#378ADD", size = 1.2) +
  geom_smooth(method = "lm", se = FALSE, color = "#A32D2D", linewidth = 0.8) +
  annotate("text", x = -Inf, y = Inf, hjust = -0.1, vjust = 1.5,
           label = t_label, size = 3.5, color = "gray30") +
  labs(title = "Theta Recovery (True vs. EAP Estimated)",
       x = "True \u03B8", y = "Estimated \u03B8 (EAP)") +
  theme_minimal(base_size = 12) +
  theme(plot.title = element_text(face = "bold", size = 13))

p_hist <- ggplot(data.frame(theta = theta_true), aes(x = theta)) +
  geom_histogram(aes(y = after_stat(density)),
                 bins = 35, fill = "#B5D4F4", color = "white") +
  stat_function(fun = dnorm, color = "#185FA5", linewidth = 1) +
  labs(title = "Simulated \u03B8 Distribution (N = 500)",
       x = "\u03B8  (lower = progressive  \u2192  higher = conservative)",
       y = "Density") +
  theme_minimal(base_size = 12) +
  theme(plot.title = element_text(face = "bold", size = 13))

# Combine into one PDF page using cowplot-style layout via ggplot
library(ggplot2)

# Save each plot separately then combine
ggsave("outputs/plot_a_recovery.pdf",    p_a,     width = 7, height = 5.5)
ggsave("outputs/plot_b_recovery.pdf",    p_b,     width = 7, height = 5.5)
ggsave("outputs/plot_theta_recovery.pdf", p_theta, width = 7, height = 5)
ggsave("outputs/plot_theta_dist.pdf",    p_hist,  width = 7, height = 4.5)

cat("\nExported: plot_a_recovery.pdf, plot_b_recovery.pdf,\n")
cat("          plot_theta_recovery.pdf, plot_theta_dist.pdf\n")

# --- Plot 2: Item Characteristic Curves (ICC) for all 30 items ---

# Build long dataframe of ICC curves for all items
icc_rows <- list()
for (j in 1:n_items) {
  p_true_vec <- p_2pl(theta_seq, item_params$a_true[j], item_params$b_true[j])
  p_est_vec  <- p_2pl(theta_seq, recovery$a_est[j],     recovery$b_est[j])
  icc_rows[[j]] <- data.frame(
    theta    = rep(theta_seq, 2),
    prob     = c(p_true_vec, p_est_vec),
    type     = rep(c("True", "Estimated"), each = length(theta_seq)),
    item_id  = j,
    item_label = paste0(j, ". ", item_params$item_name[j]),
    category = item_params$category[j],
    b_true   = item_params$b_true[j]
  )
}
icc_df <- do.call(rbind, icc_rows)
icc_df$item_label <- factor(icc_df$item_label,
                             levels = paste0(1:30, ". ", item_params$item_name))

p_icc <- ggplot(icc_df, aes(x = theta, y = prob,
                              color = category, linetype = type, linewidth = type)) +
  geom_hline(yintercept = 0.5, color = "gray80", linewidth = 0.4) +
  geom_vline(aes(xintercept = b_true), color = "gray80", linewidth = 0.4) +
  geom_line() +
  scale_color_manual(values = cat_colors, name = NULL, guide = "none") +
  scale_linetype_manual(values = c("True" = "solid", "Estimated" = "dashed"),
                        name = NULL) +
  scale_linewidth_manual(values = c("True" = 0.9, "Estimated" = 0.6),
                         name = NULL) +
  facet_wrap(~ item_label, ncol = 5) +
  scale_x_continuous(breaks = c(-2, 0, 2)) +
  scale_y_continuous(breaks = c(0, 0.5, 1), labels = c("0", ".5", "1")) +
  labs(
    title    = "Item Characteristic Curves — True (solid) vs. Estimated (dashed)",
    subtitle = "Vertical line = true b location  |  Horizontal line = P = 0.50",
    x        = "\u03B8  (lower = progressive  \u2192  higher = conservative)",
    y        = "P(Conservative response)"
  ) +
  theme_minimal(base_size = 9) +
  theme(
    plot.title    = element_text(face = "bold", size = 12),
    plot.subtitle = element_text(size = 9, color = "gray40"),
    strip.text    = element_text(size = 7.5, face = "bold"),
    legend.position = "bottom",
    panel.grid.minor = element_blank()
  )

ggsave("outputs/icc_plots.pdf", p_icc, width = 17, height = 22)
cat("Exported: icc_plots.pdf\n")

# --- Plot 3: Test Information Function ---

# Compute TIF manually from estimated parameters
tif_vals <- sapply(theta_seq, function(th) {
  sum(sapply(1:n_items, function(j) {
    a <- recovery$a_est[j]
    b <- recovery$b_est[j]
    p <- p_2pl(th, a, b)
    q <- 1 - p
    a^2 * p * q
  }))
})

tif_df <- data.frame(theta = theta_seq, information = tif_vals)

p_tif <- ggplot(tif_df, aes(x = theta, y = information)) +
  geom_line(color = "#185FA5", linewidth = 1.2) +
  geom_vline(xintercept = c(-2, 2), linetype = "dashed", color = "gray60") +
  annotate("text", x = -2.1, y = max(tif_vals) * 0.92,
           label = "-2\u03C3", hjust = 1, size = 3.2, color = "gray40") +
  annotate("text", x =  2.1, y = max(tif_vals) * 0.92,
           label = "+2\u03C3", hjust = 0, size = 3.2, color = "gray40") +
  labs(
    title    = "Test Information Function",
    subtitle = "Higher values = more precise measurement at that theta level",
    x        = "\u03B8  (lower = progressive  \u2192  higher = conservative)",
    y        = "Information"
  ) +
  theme_minimal(base_size = 12) +
  theme(
    plot.title    = element_text(face = "bold", size = 13),
    plot.subtitle = element_text(size = 10, color = "gray40")
  )

ggsave("outputs/test_information.pdf", p_tif, width = 8, height = 5)
cat("Exported: test_information.pdf\n")

# =============================================================================
# SECTION 9: Final Summary Report
# =============================================================================

cat("\n")
cat("=============================================================\n")
cat("  SIMULATION COMPLETE — OUTPUT FILES\n")
cat("=============================================================\n")
cat("  simulated_responses_500.csv   — 500x30 response matrix\n")
cat("  parameter_recovery.csv        — True vs. estimated params\n")
cat("  plot_a_recovery.pdf           — Discrimination recovery\n")
cat("  plot_b_recovery.pdf           — Item location recovery\n")
cat("  plot_theta_recovery.pdf       — Theta recovery scatter\n")
cat("  plot_theta_dist.pdf           — Theta distribution\n")
cat("  icc_plots.pdf                 — ICC for all 30 items\n")
cat("  test_information.pdf          — Test information function\n")
cat("=============================================================\n\n")

cat("RECOVERY SUMMARY:\n")
cat(sprintf("  a correlation: %.3f  (target: > 0.95)\n",
            cor(recovery$a_true, recovery$a_est)))
cat(sprintf("  b correlation: %.3f  (target: > 0.98)\n",
            cor(recovery$b_true, recovery$b_est)))
cat(sprintf("  theta r:       %.3f  (target: > 0.90)\n", theta_corr))

cat("\nFLAGGED ITEMS (|a_bias| > 0.3 or |b_bias| > 0.3):\n")
flags <- recovery[recovery$a_abias > 0.3 | recovery$b_abias > 0.3,
                  c("item_id","item_name","a_true","a_est","b_true","b_est")]
if (nrow(flags) > 0) {
  print(flags)
} else {
  cat("  None — all items recovered within tolerance.\n")
}

cat("\nNEXT STEPS:\n")
cat("  1. If a-correlation < 0.95: consider increasing N or revising\n")
cat("     extreme items (|a| > 2.0 with |b| > 1.5).\n")
cat("  2. Review flagged items for potential parameter revision.\n")
cat("  3. Check item fit (S-X2): p < .05 may indicate items that\n")
cat("     behave differently than the 2PL model predicts.\n")
cat("  4. Examine TIF: if information peaks away from center,\n")
cat("     consider adding/revising items near theta=0.\n")
cat("  5. When using real data, fix theta scale by constraining\n")
cat("     mean(theta) = 0, SD(theta) = 1 in mirt model spec.\n")


```


# Subscale Analysis

```{r}
# =============================================================================
# Economic Policy Attitudes — Subscale Score Distribution Analysis
# =============================================================================
#
# Run this AFTER irt_simulation.R — it reads simulated_responses_500.csv
# and uses the item_params object. Either source irt_simulation.R first,
# or run both scripts in the same R session.
#
# If running standalone, the item metadata is re-declared below.
# =============================================================================

# --- Packages (auto-install if needed) ---
required <- c("ggplot2", "dplyr", "tidyr")
new_pkgs <- required[!(required %in% installed.packages()[, "Package"])]
if (length(new_pkgs) > 0) install.packages(new_pkgs, repos = "https://cloud.r-project.org")

library(ggplot2)
library(dplyr)
library(tidyr)

# =============================================================================
# SECTION 1: Item Metadata (re-declared for standalone use)
# =============================================================================

item_meta <- data.frame(
  item_id  = 1:30,
  category = c(
    rep("Taxation & Wealth Inequality",     5),   # Items 1-5
    rep("Government Spending & Debt",        5),   # Items 6-10
    rep("Labor Markets & Worker Rights",     5),   # Items 11-15
    rep("Healthcare & Public Services",      5),   # Items 16-20
    rep("Corporate Regulation",              5),   # Items 21-25
    rep("Trade & Economic Nationalism",      5)    # Items 26-30
  ),
  stringsAsFactors = FALSE
)

categories <- unique(item_meta$category)

cat_colors <- c(
  "Taxation & Wealth Inequality"     = "#185FA5",
  "Government Spending & Debt"       = "#0F6E56",
  "Labor Markets & Worker Rights"    = "#854F0B",
  "Healthcare & Public Services"     = "#993556",
  "Corporate Regulation"             = "#533AB7",
  "Trade & Economic Nationalism"     = "#A32D2D"
)

# =============================================================================
# SECTION 2: Load Response Data
# =============================================================================

responses <- read.csv("simulated_responses_500.csv", row.names = 1)

# Separate item responses from theta column
item_cols  <- grep("^item_", names(responses), value = TRUE)
resp_items <- responses[, item_cols]       # 500 x 30 matrix of 0/1
theta_true <- responses$theta_true         # True theta values

cat(sprintf("Loaded: %d respondents x %d items\n", nrow(resp_items), ncol(resp_items)))

# =============================================================================
# SECTION 3: Compute Subscale Scores
# =============================================================================
# Each subscale = sum of 5 binary items (range 0-5)
# Score of 0 = all progressive responses
# Score of 5 = all conservative responses

subscale_scores <- data.frame(row.names = rownames(resp_items))

for (cat in categories) {
  cat_items <- item_meta$item_id[item_meta$category == cat]
  col_names <- paste0("item_", sprintf("%02d", cat_items))
  subscale_scores[[cat]] <- rowSums(resp_items[, col_names])
}

subscale_scores$theta_true  <- theta_true
subscale_scores$total_score <- rowSums(resp_items)

cat("\n=== Subscale Score Ranges (all should be 0-5) ===\n")
for (cat in categories) {
  cat(sprintf("  %-40s  min=%d  max=%d\n",
              cat, min(subscale_scores[[cat]]), max(subscale_scores[[cat]])))
}

# =============================================================================
# SECTION 4: Descriptive Statistics per Subscale
# =============================================================================

desc_stats <- data.frame(
  Category     = categories,
  N            = nrow(subscale_scores),
  Min          = sapply(categories, function(c) min(subscale_scores[[c]])),
  Max          = sapply(categories, function(c) max(subscale_scores[[c]])),
  Mean         = sapply(categories, function(c) round(mean(subscale_scores[[c]]), 3)),
  Median       = sapply(categories, function(c) median(subscale_scores[[c]])),
  SD           = sapply(categories, function(c) round(sd(subscale_scores[[c]]), 3)),
  Skewness     = sapply(categories, function(c) {
                   x <- subscale_scores[[c]]
                   round((mean(x) - median(x)) / sd(x), 3)  # Pearson's 2nd skewness
                 }),
  Pct_0        = sapply(categories, function(c)
                   round(mean(subscale_scores[[c]] == 0) * 100, 1)),
  Pct_5        = sapply(categories, function(c)
                   round(mean(subscale_scores[[c]] == 5) * 100, 1)),
  Cor_Theta    = sapply(categories, function(c)
                   round(cor(subscale_scores[[c]], theta_true), 3)),
  stringsAsFactors = FALSE,
  row.names    = NULL
)

cat("\n=== Subscale Descriptive Statistics ===\n")
print(desc_stats, row.names = FALSE)

# Also print total score stats
cat(sprintf(
  "\nTotal Score (all 30 items):  Mean=%.2f  SD=%.2f  Min=%d  Max=%d  r(theta)=%.3f\n",
  mean(subscale_scores$total_score), sd(subscale_scores$total_score),
  min(subscale_scores$total_score),  max(subscale_scores$total_score),
  cor(subscale_scores$total_score, theta_true)
))

# Export descriptives
write.csv(desc_stats, "subscale_descriptives.csv", row.names = FALSE)
cat("\nExported: subscale_descriptives.csv\n")

# =============================================================================
# SECTION 5: Score Frequency Tables per Subscale
# =============================================================================

cat("\n=== Score Frequency Tables (counts and %) ===\n")

freq_list <- list()
for (cat in categories) {
  tbl <- table(subscale_scores[[cat]])
  freq_df <- data.frame(
    Score    = 0:5,
    Count    = as.integer(tbl[match(0:5, names(tbl))]),
    stringsAsFactors = FALSE
  )
  freq_df$Count[is.na(freq_df$Count)] <- 0
  freq_df$Percent <- round(freq_df$Count / nrow(subscale_scores) * 100, 1)
  freq_df$Category <- cat
  freq_list[[cat]] <- freq_df

  cat(sprintf("\n  %s\n", cat))
  cat(sprintf("  %-6s %-8s %s\n", "Score", "Count", "Percent"))
  for (i in 1:nrow(freq_df)) {
    cat(sprintf("    %d     %-8d %.1f%%\n",
                freq_df$Score[i], freq_df$Count[i], freq_df$Percent[i]))
  }
}

freq_all <- do.call(rbind, freq_list)

# =============================================================================
# SECTION 6: Inter-Subscale Correlations
# =============================================================================

cat("\n=== Inter-Subscale Correlations ===\n")

score_mat  <- as.matrix(subscale_scores[, categories])
corr_matrix <- round(cor(score_mat), 3)
print(corr_matrix)

cat("\nNote: Subscales measuring the same underlying trait (economic ideology)\n")
cat("should correlate positively. Very high correlations (> 0.85) suggest\n")
cat("subscales may not be measuring distinct facets.\n")

write.csv(as.data.frame(corr_matrix), "subscale_correlations.csv")
cat("Exported: subscale_correlations.csv\n")

# =============================================================================
# SECTION 7: Plots
# =============================================================================

# --- Plot 1: Score distributions — histogram per subscale ---

long_scores <- subscale_scores %>%
  select(all_of(categories), theta_true) %>%
  pivot_longer(cols = all_of(categories),
               names_to = "Category", values_to = "Score")

p_hist <- ggplot(long_scores, aes(x = Score, fill = Category)) +
  geom_histogram(binwidth = 1, color = "white", linewidth = 0.4,
                 boundary = -0.5) +
  geom_vline(aes(xintercept = 2.5), linetype = "dashed",
             color = "gray50", linewidth = 0.5) +
  scale_fill_manual(values = cat_colors, guide = "none") +
  scale_x_continuous(breaks = 0:5) +
  facet_wrap(~ Category, ncol = 2) +
  labs(
    title    = "Subscale Score Distributions (N = 500)",
    subtitle = "Scores range 0-5  |  0 = all progressive, 5 = all conservative  |  dashed line = midpoint",
    x        = "Subscale Score",
    y        = "Count"
  ) +
  theme_minimal(base_size = 11) +
  theme(
    plot.title    = element_text(face = "bold", size = 13),
    plot.subtitle = element_text(size = 9, color = "gray40"),
    strip.text    = element_text(face = "bold", size = 9),
    panel.grid.minor = element_blank()
  )

ggsave("subscale_distributions.pdf", p_hist, width = 10, height = 10)
cat("\nExported: subscale_distributions.pdf\n")

# --- Plot 2: Subscale score vs. theta (scatter + smooth) ---

p_theta <- ggplot(long_scores, aes(x = theta_true, y = Score, color = Category)) +
  geom_point(alpha = 0.15, size = 0.9) +
  geom_smooth(method = "loess", se = TRUE, linewidth = 1.1,
              color = "gray20", fill = "gray80", alpha = 0.3) +
  scale_color_manual(values = cat_colors, guide = "none") +
  scale_y_continuous(breaks = 0:5) +
  facet_wrap(~ Category, ncol = 2) +
  labs(
    title    = "Subscale Score vs. True Theta",
    subtitle = "Points = individual respondents  |  Line = loess smooth with 95% CI",
    x        = "\u03B8  (lower = progressive  \u2192  higher = conservative)",
    y        = "Subscale Score"
  ) +
  theme_minimal(base_size = 11) +
  theme(
    plot.title    = element_text(face = "bold", size = 13),
    plot.subtitle = element_text(size = 9, color = "gray40"),
    strip.text    = element_text(face = "bold", size = 9),
    panel.grid.minor = element_blank()
  )

ggsave("subscale_vs_theta.pdf", p_theta, width = 10, height = 10)
cat("Exported: subscale_vs_theta.pdf\n")

# --- Plot 3: Correlation heatmap ---

corr_long <- as.data.frame(corr_matrix) %>%
  mutate(Cat1 = rownames(corr_matrix)) %>%
  pivot_longer(cols = -Cat1, names_to = "Cat2", values_to = "r")

# Shorten labels for readability
short_labels <- c(
  "Taxation & Wealth Inequality"     = "Taxation",
  "Government Spending & Debt"       = "Gov Spending",
  "Labor Markets & Worker Rights"    = "Labor",
  "Healthcare & Public Services"     = "Healthcare",
  "Corporate Regulation"             = "Corp Reg",
  "Trade & Economic Nationalism"     = "Trade"
)

corr_long$Cat1_short <- short_labels[corr_long$Cat1]
corr_long$Cat2_short <- short_labels[corr_long$Cat2]
corr_long$Cat1_short <- factor(corr_long$Cat1_short, levels = short_labels)
corr_long$Cat2_short <- factor(corr_long$Cat2_short, levels = short_labels)

p_corr <- ggplot(corr_long, aes(x = Cat2_short, y = Cat1_short, fill = r)) +
  geom_tile(color = "white", linewidth = 0.8) +
  geom_text(aes(label = sprintf("%.2f", r)),
            size = 3.5, color = "white", fontface = "bold") +
  scale_fill_gradient2(low = "#FCEBEB", mid = "#B5D4F4", high = "#042C53",
                       midpoint = 0.5, limits = c(0, 1),
                       name = "r") +
  labs(
    title    = "Inter-Subscale Correlations",
    subtitle = "All subscales should correlate positively if measuring the same underlying trait",
    x = NULL, y = NULL
  ) +
  theme_minimal(base_size = 11) +
  theme(
    plot.title    = element_text(face = "bold", size = 13),
    plot.subtitle = element_text(size = 9, color = "gray40"),
    axis.text.x   = element_text(angle = 30, hjust = 1),
    panel.grid    = element_blank()
  )

ggsave("subscale_correlations.pdf", p_corr, width = 8, height = 7)
cat("Exported: subscale_correlations.pdf\n")

# --- Plot 4: Mean subscale scores with SD error bars ---

summary_df <- desc_stats %>%
  mutate(
    SE    = SD / sqrt(N),
    lower = Mean - SD,
    upper = Mean + SD,
    Category_short = short_labels[Category],
    Category_short = factor(Category_short, levels = short_labels)
  )

p_means <- ggplot(summary_df,
                  aes(x = Category_short, y = Mean, color = Category)) +
  geom_hline(yintercept = 2.5, linetype = "dashed",
             color = "gray60", linewidth = 0.6) +
  geom_errorbar(aes(ymin = lower, ymax = upper),
                width = 0.2, linewidth = 0.8) +
  geom_point(size = 4) +
  scale_color_manual(values = cat_colors, guide = "none") +
  scale_y_continuous(limits = c(0, 5), breaks = 0:5) +
  annotate("text", x = 0.55, y = 2.65, label = "midpoint",
           size = 3, color = "gray50", hjust = 0) +
  labs(
    title    = "Subscale Mean Scores \u00B1 1 SD",
    subtitle = "Scores below 2.5 = net progressive lean  |  above 2.5 = net conservative lean",
    x        = NULL,
    y        = "Mean Subscale Score (0-5)"
  ) +
  theme_minimal(base_size = 11) +
  theme(
    plot.title    = element_text(face = "bold", size = 13),
    plot.subtitle = element_text(size = 9, color = "gray40"),
    axis.text.x   = element_text(angle = 20, hjust = 1),
    panel.grid.minor = element_blank()
  )

ggsave("subscale_means.pdf", p_means, width = 8, height = 5.5)
cat("Exported: subscale_means.pdf\n")

# =============================================================================
# SECTION 8: Summary
# =============================================================================

cat("\n")
cat("=============================================================\n")
cat("  SUBSCALE ANALYSIS COMPLETE — OUTPUT FILES\n")
cat("=============================================================\n")
cat("  subscale_descriptives.csv     — Mean, SD, skew, r(theta)\n")
cat("  subscale_correlations.csv     — 6x6 inter-subscale corr\n")
cat("  subscale_distributions.pdf    — Histograms per subscale\n")
cat("  subscale_vs_theta.pdf         — Score vs. theta scatter\n")
cat("  subscale_correlations.pdf     — Correlation heatmap\n")
cat("  subscale_means.pdf            — Mean ± SD per subscale\n")
cat("=============================================================\n\n")

cat("THINGS TO LOOK FOR:\n\n")
cat("  1. MEAN vs. MIDPOINT (2.5)\n")
cat("     Means well below 2.5 indicate items in that subscale\n")
cat("     sit on the progressive side of the theta continuum.\n")
cat("     This is expected for Healthcare and Labor based on b values.\n\n")
cat("  2. SD AND SPREAD\n")
cat("     Low SD (< 1.0) means most respondents gave similar answers —\n")
cat("     the subscale may not be discriminating well across the sample.\n\n")
cat("  3. FLOOR/CEILING EFFECTS\n")
cat("     High Pct_0 or Pct_5 flags subscales where the items are\n")
cat("     too extreme for your sample — consider revising b values.\n\n")
cat("  4. COR_THETA\n")
cat("     Each subscale should correlate > 0.50 with true theta.\n")
cat("     Low correlations suggest the subscale items are not well\n")
cat("     aligned with the overall construct.\n\n")
cat("  5. INTER-SUBSCALE CORRELATIONS\n")
cat("     All positive = good construct coherence.\n")
cat("     r > 0.85 between two subscales = possible redundancy.\n")
cat("     r < 0.25 = subscale may be tapping a distinct dimension.\n")


```



# subscale redo

```{r}
# =============================================================================
# Economic Policy Attitudes — Subscale Score Distribution Analysis
# =============================================================================
#
# Run this AFTER irt_simulation.R — it reads simulated_responses_500.csv
# and uses the item_params object. Either source irt_simulation.R first,
# or run both scripts in the same R session.
#
# If running standalone, the item metadata is re-declared below.
# =============================================================================

# --- Packages (auto-install if needed) ---
required <- c("ggplot2", "dplyr", "tidyr")
new_pkgs <- required[!(required %in% installed.packages()[, "Package"])]
if (length(new_pkgs) > 0) install.packages(new_pkgs, repos = "https://cloud.r-project.org")

library(ggplot2)
library(dplyr)
library(tidyr)

# =============================================================================
# SECTION 1: Item Metadata (re-declared for standalone use)
# =============================================================================

item_meta <- data.frame(
  item_id  = 1:30,
  category = c(
    rep("Taxation & Wealth Inequality",     5),   # Items 1-5
    rep("Government Spending & Debt",        5),   # Items 6-10
    rep("Labor Markets & Worker Rights",     5),   # Items 11-15
    rep("Healthcare & Public Services",      5),   # Items 16-20
    rep("Corporate Regulation",              5),   # Items 21-25
    rep("Trade & Economic Nationalism",      5)    # Items 26-30
  ),
  stringsAsFactors = FALSE
)

categories <- unique(item_meta$category)

cat_colors <- c(
  "Taxation & Wealth Inequality"     = "#185FA5",
  "Government Spending & Debt"       = "#0F6E56",
  "Labor Markets & Worker Rights"    = "#854F0B",
  "Healthcare & Public Services"     = "#993556",
  "Corporate Regulation"             = "#533AB7",
  "Trade & Economic Nationalism"     = "#A32D2D"
)

# =============================================================================
# SECTION 2: Load Response Data
# =============================================================================

responses <- read.csv("simulated_responses_500.csv", row.names = 1)

# Separate item responses from theta column
item_cols  <- grep("^item_", names(responses), value = TRUE)
resp_items <- responses[, item_cols]       # 500 x 30 matrix of 0/1
theta_true <- responses$theta_true         # True theta values

cat(sprintf("Loaded: %d respondents x %d items\n", nrow(resp_items), ncol(resp_items)))

# =============================================================================
# SECTION 3: Compute Subscale Scores
# =============================================================================
# Each subscale = sum of 5 binary items (range 0-5)
# Score of 0 = all progressive responses
# Score of 5 = all conservative responses

subscale_scores <- data.frame(row.names = rownames(resp_items))

for (cat in categories) {
  cat_items <- item_meta$item_id[item_meta$category == cat]
  col_names <- paste0("item_", sprintf("%02d", cat_items))
  subscale_scores[[cat]] <- rowSums(resp_items[, col_names])
}

subscale_scores$theta_true  <- theta_true
subscale_scores$total_score <- rowSums(resp_items)

cat("\n=== Subscale Score Ranges (all should be 0-5) ===\n")
for (cat in categories) {
  cat(sprintf("  %-40s  min=%d  max=%d\n",
              cat, min(subscale_scores[[cat]]), max(subscale_scores[[cat]])))
}

# =============================================================================
# SECTION 4: Descriptive Statistics per Subscale
# =============================================================================

desc_stats <- data.frame(
  Category     = categories,
  N            = nrow(subscale_scores),
  Min          = sapply(categories, function(c) min(subscale_scores[[c]])),
  Max          = sapply(categories, function(c) max(subscale_scores[[c]])),
  Mean         = sapply(categories, function(c) round(mean(subscale_scores[[c]]), 3)),
  Median       = sapply(categories, function(c) median(subscale_scores[[c]])),
  SD           = sapply(categories, function(c) round(sd(subscale_scores[[c]]), 3)),
  Skewness     = sapply(categories, function(c) {
                   x <- subscale_scores[[c]]
                   round((mean(x) - median(x)) / sd(x), 3)  # Pearson's 2nd skewness
                 }),
  Pct_0        = sapply(categories, function(c)
                   round(mean(subscale_scores[[c]] == 0) * 100, 1)),
  Pct_5        = sapply(categories, function(c)
                   round(mean(subscale_scores[[c]] == 5) * 100, 1)),
  Cor_Theta    = sapply(categories, function(c)
                   round(cor(subscale_scores[[c]], theta_true), 3)),
  stringsAsFactors = FALSE,
  row.names    = NULL
)

cat("\n=== Subscale Descriptive Statistics ===\n")
print(desc_stats, row.names = FALSE)

# Also print total score stats
cat(sprintf(
  "\nTotal Score (all 30 items):  Mean=%.2f  SD=%.2f  Min=%d  Max=%d  r(theta)=%.3f\n",
  mean(subscale_scores$total_score), sd(subscale_scores$total_score),
  min(subscale_scores$total_score),  max(subscale_scores$total_score),
  cor(subscale_scores$total_score, theta_true)
))

# Export descriptives
write.csv(desc_stats, "subscale_descriptives.csv", row.names = FALSE)
cat("\nExported: subscale_descriptives.csv\n")

# =============================================================================
# SECTION 5: Score Frequency Tables per Subscale
# =============================================================================

cat("\n=== Score Frequency Tables (counts and %) ===\n")

freq_list <- list()
for (cat in categories) {
  tbl <- table(subscale_scores[[cat]])
  freq_df <- data.frame(
    Score    = 0:5,
    Count    = as.integer(tbl[match(0:5, names(tbl))]),
    stringsAsFactors = FALSE
  )
  freq_df$Count[is.na(freq_df$Count)] <- 0
  freq_df$Percent <- round(freq_df$Count / nrow(subscale_scores) * 100, 1)
  freq_df$Category <- cat
  freq_list[[cat]] <- freq_df

  cat(sprintf("\n  %s\n", cat))
  cat(sprintf("  %-6s %-8s %s\n", "Score", "Count", "Percent"))
  for (i in 1:nrow(freq_df)) {
    cat(sprintf("    %d     %-8d %.1f%%\n",
                freq_df$Score[i], freq_df$Count[i], freq_df$Percent[i]))
  }
}

freq_all <- do.call(rbind, freq_list)

# =============================================================================
# SECTION 6: Inter-Subscale Correlations
# =============================================================================

cat("\n=== Inter-Subscale Correlations ===\n")

score_mat  <- as.matrix(subscale_scores[, categories])
corr_matrix <- round(cor(score_mat), 3)
print(corr_matrix)

cat("\nNote: Subscales measuring the same underlying trait (economic ideology)\n")
cat("should correlate positively. Very high correlations (> 0.85) suggest\n")
cat("subscales may not be measuring distinct facets.\n")

write.csv(as.data.frame(corr_matrix), "subscale_correlations.csv")
cat("Exported: subscale_correlations.csv\n")

# =============================================================================
# SECTION 7: Plots
# =============================================================================

# Shared label shortener used across several plots
short_labels <- c(
  "Taxation & Wealth Inequality"     = "Taxation",
  "Government Spending & Debt"       = "Gov Spending",
  "Labor Markets & Worker Rights"    = "Labor",
  "Healthcare & Public Services"     = "Healthcare",
  "Corporate Regulation"             = "Corp Reg",
  "Trade & Economic Nationalism"     = "Trade"
)

# Long-format scores for faceted plots
long_scores <- subscale_scores %>%
  select(all_of(categories), theta_true) %>%
  pivot_longer(cols = all_of(categories),
               names_to = "Category", values_to = "Score") %>%
  mutate(Category = factor(Category, levels = categories))

# --- Plot 1: Six individual labeled score distribution histograms ---
# One PDF page per subscale, each clearly titled and annotated

pdf("subscale_distributions.pdf", width = 7, height = 5)

for (cat in categories) {
  cat_data  <- subscale_scores[[cat]]
  cat_color <- cat_colors[cat]
  cat_mean  <- mean(cat_data)
  cat_sd    <- sd(cat_data)
  cat_n     <- length(cat_data)

  # Build frequency table for this subscale
  freq_tbl <- as.data.frame(table(Score = cat_data))
  freq_tbl$Score   <- as.integer(as.character(freq_tbl$Score))
  freq_tbl$Percent <- round(freq_tbl$Freq / cat_n * 100, 1)

  # Fill in any missing score values (0-5) with zero
  all_scores <- data.frame(Score = 0:5)
  freq_tbl   <- merge(all_scores, freq_tbl, by = "Score", all.x = TRUE)
  freq_tbl$Freq[is.na(freq_tbl$Freq)]       <- 0
  freq_tbl$Percent[is.na(freq_tbl$Percent)] <- 0

  # Annotation string: mean, SD, skewness, r(theta)
  r_theta  <- cor(cat_data, theta_true)
  skew_val <- (cat_mean - median(cat_data)) / cat_sd
  annot    <- sprintf(
    "Mean = %.2f   SD = %.2f\nSkewness = %.2f   r(\u03B8) = %.2f",
    cat_mean, cat_sd, skew_val, r_theta
  )

  p <- ggplot(freq_tbl, aes(x = Score, y = Freq)) +
    geom_col(fill = cat_color, color = "white", width = 0.75) +
    geom_text(aes(label = ifelse(Freq > 0, paste0(Percent, "%"), "")),
              vjust = -0.4, size = 3.2, color = "gray30") +
    geom_vline(xintercept = 2.5, linetype = "dashed",
               color = "gray50", linewidth = 0.6) +
    geom_vline(xintercept = cat_mean, linetype = "solid",
               color = cat_color, linewidth = 1.0, alpha = 0.6) +
    annotate("text", x = 2.5, y = max(freq_tbl$Freq) * 0.97,
             label = "midpoint", hjust = -0.1, size = 2.8, color = "gray50") +
    annotate("text", x = cat_mean, y = max(freq_tbl$Freq) * 0.85,
             label = sprintf("mean\n%.2f", cat_mean),
             hjust = ifelse(cat_mean < 2.5, -0.15, 1.15),
             size = 2.8, color = cat_color) +
    annotate("text", x = 5, y = max(freq_tbl$Freq) * 0.97,
             label = annot, hjust = 1, vjust = 1,
             size = 2.8, color = "gray30") +
    scale_x_continuous(breaks = 0:5,
                       labels = c("0\n(all progressive)", "1", "2", "3", "4",
                                  "5\n(all conservative)")) +
    scale_y_continuous(expand = expansion(mult = c(0, 0.15))) +
    labs(
      title    = cat,
      subtitle = sprintf("N = %d  |  Score range 0\u20135  |  Dashed = midpoint  |  Solid = mean",
                         cat_n),
      x        = "Subscale Score",
      y        = "Count"
    ) +
    theme_minimal(base_size = 12) +
    theme(
      plot.title      = element_text(face = "bold", size = 14, color = cat_color),
      plot.subtitle   = element_text(size = 9, color = "gray40"),
      panel.grid.minor = element_blank(),
      panel.grid.major.x = element_blank(),
      axis.text.x     = element_text(size = 9)
    )

  print(p)
}

dev.off()
cat("\nExported: subscale_distributions.pdf  (6 pages, one per subscale)\n")

# --- Plot 2: All six distributions on one comparison page ---

p_all <- ggplot(long_scores, aes(x = Score, fill = Category)) +
  geom_histogram(binwidth = 1, color = "white", linewidth = 0.4,
                 boundary = -0.5) +
  geom_vline(xintercept = 2.5, linetype = "dashed",
             color = "gray50", linewidth = 0.5) +
  scale_fill_manual(values = cat_colors, guide = "none") +
  scale_x_continuous(breaks = 0:5,
                     labels = c("0\n(prog)", "1", "2", "3", "4", "5\n(cons)")) +
  scale_y_continuous(expand = expansion(mult = c(0, 0.12))) +
  facet_wrap(~ Category, ncol = 2, labeller = labeller(
    Category = setNames(names(cat_colors), names(cat_colors))
  )) +
  labs(
    title    = "Subscale Score Distributions — All Subscales (N = 500)",
    subtitle = "Scores range 0\u20135  |  0 = all progressive, 5 = all conservative  |  dashed line = midpoint",
    x        = "Subscale Score",
    y        = "Count"
  ) +
  theme_minimal(base_size = 11) +
  theme(
    plot.title       = element_text(face = "bold", size = 13),
    plot.subtitle    = element_text(size = 9, color = "gray40"),
    strip.text       = element_text(face = "bold", size = 9),
    panel.grid.minor = element_blank(),
    panel.grid.major.x = element_blank()
  )

ggsave("subscale_distributions_combined.pdf", p_all, width = 10, height = 10)
cat("Exported: subscale_distributions_combined.pdf  (all 6 on one page)\n")

# --- Plot 3: Subscale Test Information Function (TIF) comparison ---
# Requires item_params with a_true / b_true AND recovered estimates.
# We compute TIF from estimated parameters (a_est, b_est) if the
# recovery object exists, otherwise fall back to true parameters.

# Re-declare item params for standalone use (mirrors irt_simulation.R)
item_params_local <- data.frame(
  item_id  = 1:30,
  category = c(
    rep("Taxation & Wealth Inequality",     5),
    rep("Government Spending & Debt",        5),
    rep("Labor Markets & Worker Rights",     5),
    rep("Healthcare & Public Services",      5),
    rep("Corporate Regulation",              5),
    rep("Trade & Economic Nationalism",      5)
  ),
  a = c(
    2.0, 1.2, 2.2, 0.6, 1.1,
    2.1, 0.5, 1.0, 1.0, 0.6,
    2.3, 1.3, 0.5, 1.1, 1.8,
    2.4, 1.2, 1.1, 0.5, 0.4,
    1.9, 1.1, 1.0, 0.6, 2.0,
    1.8, 1.2, 1.0, 0.6, 0.5
  ),
  b = c(
    -1.2,  0.1, -1.5,  0.2, -0.8,
     0.3,  0.0,  0.2,  0.1,  0.3,
    -1.0, -0.5,  0.3, -0.2, -0.7,
    -1.4, -0.4, -0.3,  0.2,  0.4,
    -0.6,  0.0,  0.3,  0.5, -0.5,
     0.5,  0.3,  0.4,  0.6,  0.2
  ),
  stringsAsFactors = FALSE
)

# If recovery object from irt_simulation.R is in memory, use estimated params
if (exists("recovery")) {
  item_params_local$a <- recovery$a_est
  item_params_local$b <- recovery$b_est
  tif_source <- "estimated (from mirt recovery)"
} else {
  tif_source <- "provisional true parameters"
}
cat(sprintf("\nTIF computed using: %s\n", tif_source))

# 2PL information function for a single item at theta
info_2pl <- function(theta, a, b) {
  p <- 1 / (1 + exp(-a * (theta - b)))
  a^2 * p * (1 - p)
}

theta_seq <- seq(-4, 4, length.out = 300)

# Build TIF dataframe: one row per (theta, subscale)
tif_rows <- list()
for (cat in categories) {
  cat_ids <- item_params_local$item_id[item_params_local$category == cat]
  tif_vals <- sapply(theta_seq, function(th) {
    sum(sapply(cat_ids, function(j) {
      info_2pl(th, item_params_local$a[j], item_params_local$b[j])
    }))
  })
  tif_rows[[cat]] <- data.frame(
    theta    = theta_seq,
    info     = tif_vals,
    Category = cat,
    stringsAsFactors = FALSE
  )
}

# Also compute full-test TIF
full_tif <- sapply(theta_seq, function(th) {
  sum(sapply(1:30, function(j) {
    info_2pl(th, item_params_local$a[j], item_params_local$b[j])
  }))
})

tif_df       <- do.call(rbind, tif_rows)
tif_df$Category <- factor(tif_df$Category, levels = categories)
full_tif_df  <- data.frame(theta = theta_seq, info = full_tif,
                            Category = "Full Test (all 30 items)")

# --- Plot 3a: Subscale TIFs overlaid on one panel ---
p_tif_overlay <- ggplot(tif_df, aes(x = theta, y = info,
                                     color = Category, linewidth = Category)) +
  geom_line() +
  scale_color_manual(values = cat_colors, name = NULL) +
  scale_linewidth_manual(values = rep(0.9, 6), guide = "none") +
  geom_vline(xintercept = 0, linetype = "dotted", color = "gray70") +
  labs(
    title    = "Subscale Test Information Functions — Overlaid",
    subtitle = sprintf("Based on %s  |  Each curve = 5 items", tif_source),
    x        = "\u03B8  (lower = progressive  \u2192  higher = conservative)",
    y        = "Information"
  ) +
  theme_minimal(base_size = 12) +
  theme(
    plot.title      = element_text(face = "bold", size = 13),
    plot.subtitle   = element_text(size = 9, color = "gray40"),
    legend.position = "bottom",
    legend.text     = element_text(size = 9),
    panel.grid.minor = element_blank()
  ) +
  guides(color = guide_legend(nrow = 2))

ggsave("tif_subscales_overlaid.pdf", p_tif_overlay, width = 9, height = 6)
cat("Exported: tif_subscales_overlaid.pdf\n")

# --- Plot 3b: Subscale TIFs in individual facets ---
p_tif_facet <- ggplot(tif_df, aes(x = theta, y = info, color = Category)) +
  geom_area(aes(fill = Category), alpha = 0.15, linewidth = 0) +
  geom_line(linewidth = 1.0) +
  geom_vline(xintercept = 0, linetype = "dotted",
             color = "gray60", linewidth = 0.5) +
  scale_color_manual(values = cat_colors, guide = "none") +
  scale_fill_manual(values  = cat_colors, guide = "none") +
  facet_wrap(~ Category, ncol = 2, scales = "free_y") +
  labs(
    title    = "Subscale Test Information Functions — Individual",
    subtitle = "Y-axis free per panel  |  Vertical dotted line = \u03B8 = 0  |  Area shading = information region",
    x        = "\u03B8  (lower = progressive  \u2192  higher = conservative)",
    y        = "Information"
  ) +
  theme_minimal(base_size = 11) +
  theme(
    plot.title       = element_text(face = "bold", size = 13),
    plot.subtitle    = element_text(size = 9, color = "gray40"),
    strip.text       = element_text(face = "bold", size = 9),
    panel.grid.minor = element_blank()
  )

ggsave("tif_subscales_faceted.pdf", p_tif_facet, width = 10, height = 10)
cat("Exported: tif_subscales_faceted.pdf\n")

# --- Plot 3c: Subscale TIFs stacked against full-test TIF ---
# Shows each subscale's contribution against the overall instrument

p_tif_full <- ggplot(tif_df, aes(x = theta, y = info, color = Category)) +
  geom_line(linewidth = 0.9, alpha = 0.8) +
  geom_line(data = full_tif_df,
            aes(x = theta, y = info),
            color = "gray20", linewidth = 1.3, linetype = "dashed",
            inherit.aes = FALSE) +
  annotate("text", x = 3.8, y = max(full_tif) * 0.97,
           label = "Full test", hjust = 1, size = 3,
           color = "gray20", fontface = "italic") +
  scale_color_manual(values = cat_colors, name = NULL) +
  geom_vline(xintercept = 0, linetype = "dotted", color = "gray70") +
  labs(
    title    = "Subscale vs. Full-Test Information Functions",
    subtitle = "Dashed = full test (30 items)  |  Colored = each 5-item subscale",
    x        = "\u03B8  (lower = progressive  \u2192  higher = conservative)",
    y        = "Information"
  ) +
  theme_minimal(base_size = 12) +
  theme(
    plot.title      = element_text(face = "bold", size = 13),
    plot.subtitle   = element_text(size = 9, color = "gray40"),
    legend.position = "bottom",
    legend.text     = element_text(size = 9),
    panel.grid.minor = element_blank()
  ) +
  guides(color = guide_legend(nrow = 2))

ggsave("tif_subscales_vs_fulltest.pdf", p_tif_full, width = 9, height = 6)
cat("Exported: tif_subscales_vs_fulltest.pdf\n")

# --- Plot 4: Correlation heatmap ---

corr_long <- as.data.frame(corr_matrix) %>%
  mutate(Cat1 = rownames(corr_matrix)) %>%
  pivot_longer(cols = -Cat1, names_to = "Cat2", values_to = "r")

corr_long$Cat1_short <- short_labels[corr_long$Cat1]
corr_long$Cat2_short <- short_labels[corr_long$Cat2]
corr_long$Cat1_short <- factor(corr_long$Cat1_short, levels = short_labels)
corr_long$Cat2_short <- factor(corr_long$Cat2_short, levels = short_labels)

p_corr <- ggplot(corr_long, aes(x = Cat2_short, y = Cat1_short, fill = r)) +
  geom_tile(color = "white", linewidth = 0.8) +
  geom_text(aes(label = sprintf("%.2f", r)),
            size = 3.5, color = "white", fontface = "bold") +
  scale_fill_gradient2(low = "#FCEBEB", mid = "#B5D4F4", high = "#042C53",
                       midpoint = 0.5, limits = c(0, 1), name = "r") +
  labs(
    title    = "Inter-Subscale Correlations",
    subtitle = "All subscales should correlate positively if measuring the same underlying trait",
    x = NULL, y = NULL
  ) +
  theme_minimal(base_size = 11) +
  theme(
    plot.title    = element_text(face = "bold", size = 13),
    plot.subtitle = element_text(size = 9, color = "gray40"),
    axis.text.x   = element_text(angle = 30, hjust = 1),
    panel.grid    = element_blank()
  )

ggsave("subscale_correlations.pdf", p_corr, width = 8, height = 7)
cat("Exported: subscale_correlations.pdf\n")

# --- Plot 5: Mean subscale scores with SD error bars ---

summary_df <- desc_stats %>%
  mutate(
    lower          = Mean - SD,
    upper          = Mean + SD,
    Category_short = short_labels[Category],
    Category_short = factor(Category_short, levels = short_labels)
  )

p_means <- ggplot(summary_df,
                  aes(x = Category_short, y = Mean, color = Category)) +
  geom_hline(yintercept = 2.5, linetype = "dashed",
             color = "gray60", linewidth = 0.6) +
  geom_errorbar(aes(ymin = lower, ymax = upper),
                width = 0.2, linewidth = 0.8) +
  geom_point(size = 4) +
  scale_color_manual(values = cat_colors, guide = "none") +
  scale_y_continuous(limits = c(0, 5), breaks = 0:5) +
  annotate("text", x = 0.55, y = 2.65, label = "midpoint",
           size = 3, color = "gray50", hjust = 0) +
  labs(
    title    = "Subscale Mean Scores \u00B1 1 SD",
    subtitle = "Below 2.5 = net progressive lean  |  above 2.5 = net conservative lean",
    x        = NULL,
    y        = "Mean Subscale Score (0\u20135)"
  ) +
  theme_minimal(base_size = 11) +
  theme(
    plot.title       = element_text(face = "bold", size = 13),
    plot.subtitle    = element_text(size = 9, color = "gray40"),
    axis.text.x      = element_text(angle = 20, hjust = 1),
    panel.grid.minor = element_blank()
  )

ggsave("subscale_means.pdf", p_means, width = 8, height = 5.5)
cat("Exported: subscale_means.pdf\n")

# =============================================================================
# SECTION 8: Summary
# =============================================================================

cat("\n")
cat("=============================================================\n")
cat("  SUBSCALE ANALYSIS COMPLETE — OUTPUT FILES\n")
cat("=============================================================\n")
cat("  subscale_descriptives.csv          — Mean, SD, skew, r(theta)\n")
cat("  subscale_correlations.csv          — 6x6 inter-subscale corr matrix\n")
cat("  subscale_distributions.pdf         — 6 individual labeled histograms\n")
cat("  subscale_distributions_combined.pdf— All 6 on one comparison page\n")
cat("  tif_subscales_overlaid.pdf         — All 6 TIFs on one panel\n")
cat("  tif_subscales_faceted.pdf          — Individual TIF per subscale\n")
cat("  tif_subscales_vs_fulltest.pdf      — Subscale TIFs vs. full test\n")
cat("  subscale_correlations.pdf          — Correlation heatmap\n")
cat("  subscale_means.pdf                 — Mean ± SD per subscale\n")
cat("=============================================================\n\n")

cat("THINGS TO LOOK FOR:\n\n")
cat("  1. MEAN vs. MIDPOINT (2.5)\n")
cat("     Means well below 2.5 indicate items in that subscale\n")
cat("     sit on the progressive side of the theta continuum.\n")
cat("     This is expected for Healthcare and Labor based on b values.\n\n")
cat("  2. SD AND SPREAD\n")
cat("     Low SD (< 1.0) means most respondents gave similar answers —\n")
cat("     the subscale may not be discriminating well across the sample.\n\n")
cat("  3. FLOOR/CEILING EFFECTS\n")
cat("     High Pct_0 or Pct_5 flags subscales where the items are\n")
cat("     too extreme for your sample — consider revising b values.\n\n")
cat("  4. COR_THETA\n")
cat("     Each subscale should correlate > 0.50 with true theta.\n")
cat("     Low correlations suggest the subscale items are not well\n")
cat("     aligned with the overall construct.\n\n")
cat("  5. INTER-SUBSCALE CORRELATIONS\n")
cat("     All positive = good construct coherence.\n")
cat("     r > 0.85 between two subscales = possible redundancy.\n")
cat("     r < 0.25 = subscale may be tapping a distinct dimension.\n")
```